\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-04-27 Latex/main.tex
paper updates

We focus on continuous-time models defined as a solution of a stochastic differential equation with time-varying coefficients. One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}. It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. The following formulation of the equations represents an extension of the CVM \cite{johnson_continuoustime_2008}, which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-\begin{pmatrix} 
			\frac{1}{\tau} & -\omega \\
			\omega & \frac{1}{\tau}
		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. Here, $W(t)$ is a two-dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates. The case $\omega=0$ corresponds to a standard CVM. In most applications, this model is satisfactory, but in case of birds movement, or merely tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.

@@

In general, the motion of marine mammals is subject to spatial constraints induced by the coastline. This is particularly the case for the narwhal data we want to analyse: the animals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland). Thus, constraints for the shore are needed in the location process $X$. However, this is not the case in model~\ref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}. 
Recently, a reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
In this study, simulation of constrained paths is achieved by projecting on the coastline the  points that eventually reach land. Estimation of the SDE parameters then relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data. Another idea is to add a repulsive potential function in the drift, which is done to constrain ants movement within a box in \cite{russell_spatially_2018}. In our case study, the narwhals move in a complex fjords system, which means that this technique might not be easily replicated because it relies on using very basic boundaries to derive an explicit expression for the potential function. As a consequence, there is a need to develop new models adapted to such spatial constraints.\\
Additionally, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}. For narwhals, it was observed from the data that they have a proclivity to move along the shoreline. Here we propose a new model based on the following idea: as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline. The novelty of this paper consists in modeling the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector. This makes the drift in the velocity equation dependent on the location process $X$ and the boundary of the domain.  \\ %More biblio about estimation methods
We show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}. To make this possible, we formulate a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}. Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\

@@

Section 2 gives an overview of the narwhal data available for the analysis. Then in Section 3, we discuss the diffusion models and the introduction of covariates in the diffusion parameters through mixed effect models. Section 4 is dedicated to a comprehensive simulation study, whereas in the last section, the results of the application to the narwhal data are examined.

\section{Movement data of Greenland endemic narwhals}
%describe how the data was obtained, the geography, type of whales, year, exposure to boats etc..
%trajectory plots, coordinate system, trial, intertrial; gunsize, tagtime

@@

\subsection{Definition of the covariates}

@@

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain in which the narwhals can move. The closest point on the shoreline to the observed position $y_{ij}$  is denoted $p_{ij}$, and the distance to this point, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$. The resulting values range from $0$ to $7.5$ km. About $9\%$ of the GPS measurements in the dataset turned out to be on land. This is primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. For constrained movement models, these points are ignored. Then, exposure to the shore is defined with the following formula
\begin{equation}
\left\{
\begin{array}{ll}
	 E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{ref} \\
	E^{shore}_{ij}=\frac{1}{D_{shore}(t_j)} & \mbox{if } D^{shore}_{ij} \leq D_{ref}
\end{array}
\right.
\label{eq: exp shore definition}
\end{equation}
where the reference threshold $D_{ref}$ is chosen equal to $0.5$ km. We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
The angle between the narwhal's heading and the shoreline, denoted $\Theta_{ij}$, is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-p_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore. 
The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}.

@@

The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation.

\subsection{Standard Rotational correlated velocity model (RCVM)}
\label{section: RCVM}

%Discussion, different options

A general formulation for a rotational correlated velocity model is \cite{albertsen_generalizing_2018}:

\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(V(t)-\mu)dt+\Sigma dW(t)

	\end{array}
	\right.
	\label{eq: general RACVM equation}
\end{equation}
where $A=\begin{pmatrix} 
	\frac{1}{\tau_1} & -\omega \\
	\omega & \frac{1}{\tau_2}
\end{pmatrix}$ and $\Sigma$ is a lower triangular matrix with positive diagonal elements whereas $\tau_1$ and $\tau_2$ are two autocorrelations time scale for each component of the velocity.
There is an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}.

It is commonly assumed that the movement is isotropic, forcing $\tau_1=\tau_2=\tau$, and that the random changes in both components of the velocity are independent with the same variance, in which case $\Sigma=\sigma^2 I_2$ \cite{johnson_continuoustime_2008}.
The diffusion parameter $\sigma$ is written as $\sigma=\frac{2\nu}{\sqrt{\tau \pi}}$, where $\nu$ is homogeneous to a velocity. This allows easier biological interpretation  of the parameters in the equation \cite{gurarie_correlated_2017}. These simplifications produce the equation~\ref{eq: RACVM equation}.

When parameters $\tau$, $\mu$ and $\nu$ are constant over time, both the velocity and the position processes can be solved explicitly.
The transition density for the velocity process is such that 
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a rotation matrix of angle $-\omega \Delta$.
Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.

%stationary distribution, distribution of the norm, distribution of the phase
The stationary distribution of the velocity process is $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$. This result can be used to get the long term distribution of the velocity norm and the velocity phase angle. After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$. In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$. Hence, the order of magnitude of the parameter $\nu$ is similar to the overall mean empirical velocity norm. The distribution of the phase angle $\Phi$ of the velocity has been derived in \cite{pawula_distribution_1982}.
When $\mu=0$, this distribution is uniform over $[-\pi,\pi]$. Hence, taking a look at the histograms of the phase angle of the data can help to assess whether a non zero mean velocity is necessary in the model or not.

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/tau5_nu5_omega1mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/tau5_nu1_omega0mu10_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu1omega0_mu10_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/theta_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/rcvm/theta_histo_tau5_nu1omega0_mu10_sigmaobs0.03 .png}
	\caption{}
\end{subfigure}
	\caption{(a)-(b) Two samples of a RCVM with respective parameters $\tau=5 h$, $\nu=5$ km/h,$\omega=1$ rad/h, $\mu=(1,0)$ km/h  and $\tau=5 h$, $\nu=1$ km/h,$\omega=0$ rad/s, $\mu=(0,0)$ km/h. The black cross indicates the starting point of the process. (c) - (d) Histograms of the empirical velocities obtained from the samples, along with the theoretical probability density function. (e) - (f) Histograms of the empirical velocity phase angles obtained from the samples, along with the theoretical probability density function. }
	\label{fig: sample RACVM}
\end{figure}

To get a constrained version of the CVM, it is natural to make use of a RCVM when the process is close to the shore and the velocity is pointing toward the shore to represent how the animals turn in reaction to the shore.
This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
Denote $\mathcal{D} \subset \R^2$ the domain of the process $X$.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))v(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

Here, $A(X(t),V(t))=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega(D_{shore}(t),\Theta(t)) \\
	\omega(D_{shore}(t),\Theta(t)) & \frac{1}{\tau}
\end{pmatrix}$
with $D_{shore}(t)=d(X(t),\mathcal{D})$,
 $\Theta(t)=\widehat{(X(t)-p(t),V(t))}$ and $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$.

 Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
 For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
 \[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
 In this case, the formulas of section~\ref{section: RCVM} are still valid.
 The question still hanging is : what are the conditions on the function $\omega(D_{shore},\Theta)$ to ensure that the process is constrained ?
 We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $x$.
 An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation. For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$. Moreover, a rotation needs to be considered only for some specific values of the angle $\Theta$, more precisely $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$, that is when the narwhal is heading in the direction of the shore.
 The closer we are to the shore, the faster the velocity should rotate. 
 whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
 A weight based on the distance to the shore can be defined as follows :
 \[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

 Based on this weight, a relationship between $\omega$ and the angle $\Theta$ and the distance to the shore $D_{shore}$ can be formulated as follows :

 \[
 \omega(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D_{shore})
 \]
 for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.

>>>>

We focus on continuous-time models defined as a solution of a stochastic differential equation with time-varying coefficients. One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}. It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. The following formulation of the equations represents an extension of the CVM \cite{johnson_continuoustime_2008}, which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-\begin{pmatrix} 
			\frac{1}{\tau} & -\omega \\
			\omega & \frac{1}{\tau}
		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $V(t)$ is the horizontal two-dimensional velocity at time $t$ and $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates. Here, $W(t)$ is a two-dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, while $\omega$ is an angular velocity that controls how fast the velocity vector rotates. The case $\omega=0$ corresponds to a standard CVM. In most applications, this model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}.

@@

In general, the motion of marine mammals is subject to spatial constraints induced by the coastline. This is particularly the case for the narwhal data we want to analyse: the animals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland). Thus, constraints for the shore are needed in the location process $X$. However, this is not the case in model~\ref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}. 
Recently, a reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
In this study, simulation of constrained paths is achieved by projecting on the coastline the  points that eventually reach land. Estimation of the SDE parameters then relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data. Another idea is to add a repulsive potential function in the drift, which is done to constrain ants movement within a box in \cite{russell_spatially_2018}. In our case study, the narwhals move in a complex fjords system, which means that this technique might not be easily replicated because it relies on using very basic boundaries to derive an explicit expression for the potential function. As a consequence, there is a need to develop new models adapted to such spatial constraints.\\
Additionally, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}. For narwhals, it was observed from the data that they have a proclivity to move along the shoreline. Here we propose a new model based on the following idea: as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline. The novelty of this paper consists in modeling the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector. This makes the drift in the velocity equation dependent on the location process $X$ and the boundary of the domain.  \\ %More biblio about estimation methods

Here, we show that the angular velocity  can be estimated as a function of the two covariates in the framework of the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021}. To make this possible, we formulate a state-space model by binding the velocity and location in one single hidden state vector, following \cite{johnson_continuoustime_2008}. Maximum likelihood estimation based on Kalman filter is then performed as for any other model in \texttt{smoothSDE}.\\

@@

The diffusion models are discussed in Section 2
Then,  Section 3 is dedicated to a comprehensive simulation study, whereas in the last section,we give an overview of the narwhal data available for the analysis and the results of the application of our models to the narwhal data.

The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation. First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement.

\subsection{Standard Rotational Advective correlated velocity model (RACVM)}
\label{section: RACVM}

\subsubsection{Solution of the SDE}

We write $A=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega \\
	\omega & \frac{1}{\tau}
\end{pmatrix}$ and recall equation~\ref{eq: RACVM equation}

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. Then the Euler-scheme is used to get the transitions of the location process. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, both the velocity and the position processes have a closed form formula. Though this model is discussed in \cite{gurarie_correlated_2017}, we found no mention of these explicit formulas. \\

The transition density for the velocity process is such that 
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

where $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
Intuitively, the next velocity is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $v(t)$ rotated by an angle $-\omega \Delta$.

%stationary distribution, distribution of the norm, distribution of the phase
This velocity process has stationary distribution $\mathcal{N}\left(\mu,\frac{2\nu^2}{\pi} I_2\right)$ \cite{gurarie_correlated_2017}. After the velocity process has reached stationarity, that is typically for $t \geq 3\tau$, $\vert \vert v(t) \vert \vert$ follows a Rice distribution with mean parameter $\vert \vert \mu \vert \vert$ and variance parameter $\frac{2\nu^2}{\pi}$. In the special case $\mu=(0,0)$, this means that $\E(\vert \vert v(t) \vert \vert)=\nu$ and motivates the interpretation of $\nu$ as the mean velocity norm of the movement.

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/tau5_nu5_omega1mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
	\caption{(a) Samples of a RACVM with parameters $\tau=5 h$, $\nu=5$ km/h,$\omega=1$ rad/h, $\mu=(1,0)$ km/h. The black cross indicates the starting point of the process. The samples were obtained from the exact transition densities. (b) Histogram of the empirical velocities obtained from the samples, along with the theoretical probability density function.}
	\label{fig: sample RACVM}
\end{figure}

\subsubsection{State-space model}
In most applications, it is needed to consider measurement errors through a state-space model. We suppose that positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$. $Y_j$ is the random vector representing the observed position at time $t_j$, while $X_j=X(t_j)$ is the true position at time $t_j$.

\begin{equation}  \left\{
	\begin{array}{l}
		Y_j=X_j+\varepsilon_j \quad j \in \{1,\cdots,n\}\\
		\varepsilon_j \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX(t)=V(t)dt  \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
\end{equation}

Here, we consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time.
We make the approximation that these parameters are constant on each time step $[t_j,t_{j+1}]$ . Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $V_j=V(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix} 
	\frac{1}{\tau_j} & -\omega_j \\
	\omega_j & \frac{1}{\tau_j}
\end{pmatrix}$.

\begin{equation}
	\forall j \in \{1, \cdots, n\} \quad 
	\left\{ \begin{array}{l}
		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
	\end{array}\right. \\
	\label{eq: RCVM autoregressive formula}
\end{equation}

where $\zeta_j$ and $\xi_j$ are centered normal random vectors.
Denote $\alpha_j=\begin{pmatrix} X_{j,1}  & X_{j,2} & V_{j,1} & V_{j,2}\end{pmatrix}^\top$. The state space equations are

\[
\left\{
\begin{array}{l}
	Y_j=Z\alpha_j+\varepsilon_j \\
	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
\end{array}
\right.\]
where the link matrices are 
\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
\]

One can check that for $\omega_j=0$, the formulas match the ones of the CTCRW model \cite{johnson_continuoustime_2008}.
This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter. This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} the special case $\omega=0$. Therefore, we extend this inference method to the more general model~\ref{eq: RACVM equation}.

Unconstrained movement of animals that do not exhibit a clear rotational movement, as it is often the case for marine mammals, is generally modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$.
To get a constrained version of the CVM, it is natural to make use of a RCVM when the process is close to the shore and the velocity is pointing toward the shore to represent how the animals turn in reaction to the shore.
This can be achieved by choosing a smooth function that is zero far away from the shore and non zero close to the shore for $\omega$.
Denote $\mathcal{D} \subset \R^2$ the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

Here, $A(X(t),V(t))=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega(t) \\
	\omega(t) & \frac{1}{\tau}
\end{pmatrix}$
where $\omega(t)=f_{\omega}(D_{shore}(t),\Theta(t))$, $D_{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
In this case, the formulas of section~\ref{section: RACVM} are still valid.
The question still hanging is : what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation. For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$. We hint that the condition should also include the regularity of the domain, but this will be a subject for further work. Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D_{shore}$ approached $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate. 
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
A weight based on the distance to the shore can be defined as follows :
\[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D_{shore}$ can be formulated as follows :

\begin{equation}
f_{\omega}(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D_{shore})
\label{eq: smooth omega}
\end{equation}
for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. The general model is very flexible, and different species-specific behaviours could be described. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. \\
Figure~\ref{fig: trueomega} shows a smooth function obtained from~\ref{eq: smooth omega} for the parameter $\omega$. Figure~\ref{fig: splineomega} shows three spline basis functions and a smooth function $f_{\omega}$ obtained as a combination of these basis functions.

\begin{figure}[H]
	\centering
		\centering
		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
			(a) $D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
		\label{fig: trueomega}
\end{figure}

\begin{figure}[H]
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
		\caption{}
	\end{subfigure}
	\caption{}
	\label{fig: splineomega}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.4]{images/simulation study/illustrative_sample_rect.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.4]{images/simulation study/illustrative_sample_fjords.png}
		\caption{}
	\end{subfigure}
	\caption{Samples of the CRCVM in different domains. x and y axis are in metres and initial positions are indicated by a red cross and red point represent the nearest points on the boundary of the domain (a) One sample three day sample with time step $h=1$ min in a rectangular domain (b) One sample three day sample with time step $h=1$ min in Scoresby Sound fjords system in South-East Greenland}
	\label{fig: samples_rect_CRCVM}
\end{figure}

\section{Simulation study}

\label{section: simulation study}
%samples and estimation
%discussion about tensor splines vs additive splines

\section{Application to narwhals behavioral response}

@@

\subsection{Notations and definition of the covariates}

@@

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed position $y_{ij}$  is denoted $p_{ij}$, and the distance to this point, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$. The resulting values range from $0$ to $7.5$ km. About $9\%$ of the GPS measurements in the dataset turned out to be on land. This is primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. For constrained movement models, these points are ignored. Then, exposure to the shore is defined with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{ref} \\
		E^{shore}_{ij}=\frac{1}{D_{shore}(t_j)} & \mbox{if } D^{shore}_{ij} \leq D_{ref}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference threshold $D_{ref}$ is chosen equal to $0.5$ km. We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
The angle between the narwhal's heading and the shoreline, denoted $\Theta_{ij}$, is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-p_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore. 
The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
Histograms of the distance to the ship and the distance to the shore are shown in figure~\ref{fig: distance to shore and distance to ship histograms}.

@@
